\section{\genetrace}
\label{sec:genetrace}

\genetrace{} is a four-level annotation corpus over a contamination-guarded
paper pool. Section~\ref{sec:genetrace:schema} defines the schema,
Section~\ref{sec:genetrace:construction} describes the construction
pipeline, and Section~\ref{sec:genetrace:contamination} details the
contamination guard that distinguishes our release from prior paper-graph
corpora.

\subsection{Schema}
\label{sec:genetrace:schema}

\paragraph{Level 1 --- GenomeCard.} For each paper $p$, a JSON record
containing the six gene fields from \ideaevolving{} (mechanism, niche,
observation, limitation, delta, claim), each accompanied by zero or more
\emph{evidence quotes} from the source text. Evidence quotes are
character-offset-anchored and verified by exact substring match against
\texttt{title + abstract + intro}. The full schema is given in
Appendix~\ref{app:schema:card}.

\paragraph{Level 2 --- DynamicsEdge.} For each ordered pair $(p, q)$ in a
lineage chain, a record giving the dynamics label
$d \in \{\text{Mutation}, \text{Adaptive Radiation},
\text{Hybridization}, \text{Speciation}, \text{Niche Competition}\}$, the
\emph{driver} field (which of $p$'s six fields motivated the transition),
and per-field \emph{gene fates}
$\in \{\text{INHERITED}, \text{MUTATED}, \text{LOST}, \text{NOVEL},
\text{HYBRIDIZED}\}$. Edges include a teacher chain-of-thought trace
explaining the dynamics assignment.

\paragraph{Level 3 --- LineageChain.} For each chain of length $\geq 3$, a
record giving the member papers, the per-step dynamics sequence, and a
teacher narrative summary. Chains are constructed by DFS over the
DynamicsEdge graph with cycle detection.

\paragraph{Level 4 --- VerifierBundle.} Per-annotation automated scores
(schema validity, evidence groundedness, dynamics consistency,
fate--dynamics agreement) bundled in a separate file for downstream
consumers who want to filter by quality or use the scores as RL rewards.

\subsection{Construction pipeline}
\label{sec:genetrace:construction}

\paragraph{Paper pool.} We start from $\sim$3,846 pre-2017 papers extracted
from the \ideaevolving{} \texttt{paper\_db}, filtered to those with
abstracts of $\geq$200 words (yielding $\sim$857 in the round-1 SFT pool;
expanded to 5K for the v0.1 release). Pre-2017 papers and the
\ideaevolving{} denylist (Section~\ref{sec:genetrace:contamination}) jointly
guarantee disjointness from \geneexam{} evaluation papers.

\paragraph{Teacher annotation.} For each paper, we prompt GPT-5.5 (Azure
keyless endpoint, frozen snapshot \texttt{2024-12-01-preview}) with the
source text and a structured \emph{gene\_card\_extract} prompt. The teacher
returns a JSON object with the six gene fields, each accompanied by 1--3
verbatim evidence quotes from the source. We accept the annotation iff a
deterministic verifier (Appendix~\ref{app:verifier}) returns
\texttt{schema\_valid} = 1.0 and \texttt{evidence\_grounded} $\geq 0.8$.
For round-1, 100\% of teacher outputs passed the schema check; 856 of 857
passed the evidence check.

\paragraph{Lineage edges.} For each pair $(p, q)$ where $q$ cites $p$ and
both are in the safe pool, we prompt the teacher with both cards and the
candidate dynamics enum. Dynamics validity (label $\in$ enum) and
fate--dynamics consistency (e.g., \textit{Hybridization} requires $\geq 1$
\texttt{HYBRIDIZED} fate) are enforced by the verifier.

\paragraph{Quality filtering.} We discard annotations failing any verifier
check. For the v0.2 release, we additionally apply \emph{consistency
filtering} \citep{west2023symbolic}: each lineage edge is annotated twice
(once by GPT-5.5, once by Qwen3-14B-Thinking), and only edges where both
teachers agree on the dynamics label are retained.

\subsection{Contamination guard}
\label{sec:genetrace:contamination}

A paper-graph corpus intended for training models that will be evaluated on
benchmarks derived from the same literature must explicitly demonstrate
non-overlap with the evaluation set. We adopt three layers:

\begin{enumerate}
  \item \textbf{Denylist.} An explicit blocklist of 8,359 paper IDs
    extracted from \ideaevolving's \texttt{paper\_db}; 90.4\% have a resolvable
    Semantic Scholar ID. We additionally include a 1-hop OpenAlex
    citation expansion (in progress; v0.2 release).
  \item \textbf{Temporal cut.} The training pool is restricted to
    pre-2017 papers. \geneexam{} draws predominantly from 2018+ literature,
    providing a temporal disjointness guarantee independent of the
    denylist.
  \item \textbf{Min-K\%++ leakage diagnostic.} We measure
    \citep{shi2024minkpp}-style memorisation probability that the trained
    model has seen each \geneexam{} paper. We report this number in the
    dataset card; target Min-20\%++ $<$ 0.05 (consistent with no leakage).
\end{enumerate}

The denylist, safe-pool construction code, and Min-K\%++ measurement
script ship with the release, enabling any downstream consumer to verify
the contamination guard end-to-end.

\subsection{Dataset statistics}
\label{sec:genetrace:stats}

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lrrrr}
    \toprule
    Release & Cards & Edges & Chains & Status \\
    \midrule
    v0.1 (paper) & \todo{5K} & \todo{2K} & \todo{200} & in progress \\
    v0.2 (1mo)   & 20K & 8K & 1K & planned \\
    v1.0         & 50K+ & 25K+ & 5K+ & long-term \\
    \bottomrule
  \end{tabular}
  \caption{\genetrace{} release tiers. v0.1 ships with the paper.}
  \label{tab:release_tiers}
\end{table}

Table~\ref{tab:release_tiers} summarises the tiered release. The v0.1
corpus shipped with this paper contains \todo{5K} GenomeCards
($\sim$\todo{4.3K} with full evidence quotes), \todo{2K} DynamicsEdges
spanning all five modes, and \todo{200} LineageChains of length 3--5.
All annotations include teacher reasoning traces (mean length
\todo{N} tokens, capped at 1024).
